---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Using covariates

```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
             "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)
```

This vignette assumes you have a series of covariates assumed to explain a given dataset. 
To illustrate how the package works, we define arbitrarily the following covariates.

```{r setup}
library(ecodecision)
library(spatstat, quietly = TRUE)

remove(list = ls())

set.seed(1)

temperature <- function(x, y) 1 - 4 * (x - 0.5)^2
wetness <- function(x, y) 0.5 * (1 + sin(pi * 2 * (y - 1)))

covariates <- list(t = wetness)
```

Finally, we need a dataset to work on.
In order to properly illustrate our method, we want to use data which is biased according to our given covariates, as given below.
In general, this part is not necessary -- just use your own data.

```{r}
configuration <- ppjsdm::rppp(lambda = 50)
```

The function `gibbsm` fits a multivariate Gibbs point process to our dataset.
For example,

```{r}
ppjsdm::gibbsm(configuration, covariates = covariates)
```

