---
title: "Details on the model and how it can be customised"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
             "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_check,
  purl = !is_check
)
```

```{r setup}
library(ppjsdm)
library(spatstat)

remove(list = ls())

set.seed(1)
```

In this vignette, we illustrate the model with some examples.
The package includes a function to compute the Papangelou conditional intensity above, which will make things easier!

# No interaction: Poisson point process

We begin by considering a single species, with no repulsion, no covariates, and a given log-intensity.

```{r}
beta0 <- 4
model <- "Geyer"
```

Take a single sample from the corresponding point process,

```{r}
configuration <- rgibbs(beta0 = beta0, model = model)
```

and plot it below.

```{r, fig.align = 'center', fig}
par(mar = c(5, 4, 4, 13) + 0.1)
plot(configuration)
```

Now, we can compute the Papangelou conditional intensity over the region.

```{r}
x_axis <- seq(from = 0, to = 1, by = 0.001)
y_axis <- x_axis
z <- outer(x_axis, y_axis, function(x, y) {
  compute_papangelou(configuration = configuration, 
                     x = x, 
                     y = y, 
                     beta0 = beta0, 
                     model = model)
})
```

In this case, the Papangelou conditional intensity is equal to the intensity over the whole region:

```{r}
all(z == exp(beta0))
```

This means that the probability of finding a point anywhere on the region conditional on the given locations is independent of the locations, and proportional to the intensity everywhere.
In other words, the associated point process is a Poisson point process of intensity `exp(beta0)`.

# Adding interactions

We now add short-range interactions between individuals, as controlled by the parameter `alpha`.

```{r}
alpha <- -0.1
```

Let us see how this changes the sample and its Papangelou conditional intensity.
The package contains a function to sample from the distribution and plot the configuration and the Papangelou conditional intensity.
We can plot the Papangelou conditional intensity by a call to that function.
The circles represent a sample from the target distribution, and we have added the Papangelou conditional intensity as an overlay.

```{r}
plot_papangelou(alpha = alpha, beta0 = beta0, model = model)
```

The plot indicates repulsion within a given interaction radius: at locations far from existing individuals in the configuration, the Papangelou conditional intensity is equal to `exp(beta0)`, and it decreases within the interaction radius.
The parameter `alpha` controls the strength of the repulsion, and when it is increased (in absolute value) the point process becomes more repulsive.

```{r}
alpha <- -4
plot_papangelou(alpha = alpha, beta0 = beta0, model = model)
```

Making `alpha` positive instead of negative changes the behaviour.

```{r}
alpha <- 0.1
plot_papangelou(alpha = alpha, beta0 = beta0, model = model)
```

The Papangelou conditional intensity is now larger at locations closer to existing individuals, indicating that the point process is attractive.
Thus, our model is either attractive or repulsive depending on the sign of `alpha`.

The model also incorporates medium-ranges interaction via the parameter $\gamma$.
```{r}
gamma <- -0.1
medium_range <- 0.1
long_range <- 0.2
medium_range_model <- "Geyer"
plot_papangelou(alpha = alpha, 
                beta0 = beta0, 
                model = model, 
                gamma = gamma, 
                medium_range = medium_range,
                long_range = long_range,
                medium_range_model = medium_range_model)
```

# Changing features of the interaction

Different features of the interaction can be customised.
To begin with, you can control the radii of interaction (which correspond to $r_{i,j}$ in the model).

```{r}
alpha <- -0.1
short_range <- 0.05
plot_papangelou(alpha = alpha, beta0 = beta0, model = model, short_range = short_range)
```

There are also various kinds of smoothing functions available via the `model` parameter (which correspond to $\varphi_{r_{i,j}}$ in the model).
Here is what the square exponential smoothing looks like.

```{r}
model <- "square_exponential"
plot_papangelou(alpha = alpha, beta0 = beta0, model = model)
```

# Discussion of the different variations

These different variations of the model generate quite different samples in extreme cases.
For example, for a very large value of the repulsion parameter, the Geyer model exhibits extreme regularity.
The Geyer model does not have any smoothing, i.e. it corresponds to the choice $\varphi_{r_{i,j}}(r)=\mathbf1_{\{r\le r_{i,j}\}}$, where $\mathbf1$ is the indicator function.

```{r}
extreme_configuration <- rgibbs(alpha = matrix(-Inf),
                                beta0 = log(100),
                                model = "Geyer")
par(mar = c(5, 4, 4, 13) + 0.1)
plot(extreme_configuration)
```

Other variants of the model show some interesting features.
As an illustration, here is a plot of an extremely attractive square exponential version of the model.

```{r}
extreme_configuration <- rgibbs(alpha = 2.5,
                                beta0 = 0,
                                model = "square_exponential",
                                steps = 1000000)
par(mar = c(5, 4, 4, 13) + 0.1)
plot(extreme_configuration)
```

# Adding covariates

Adding covariates is also possible.
It is best to provide the covariates as `spatstat::im` objects, and it is usually easy to convert to that format thanks to the conversion functions provided by `spatstat`.

```{r}
model <- "Geyer"
covariates <- list(temperature = spatstat.geom::as.im(function(x, y) x - 0.5, W = owin()))
beta <- 2
plot_papangelou(model = model,
                alpha = alpha,
                beta0 = beta0,
                beta = beta,
                covariates = covariates)
```

<!-- # Multiple species setting -->

<!-- The package is in fact meant to work in a multiple species setting, so let us give an example in that case. -->

<!-- ```{r} -->
<!-- beta0 <- c(4, 4) -->
<!-- alpha <- cbind(c(-1, 0.1), c(0.1, -0.5)) -->
<!-- gamma <- cbind(c(-0.5, 0), c(0, -0.5)) -->
<!-- model <- "square_bump" -->
<!-- medium_range_model <- "square_exponential" -->
<!-- short_range <- cbind(c(0.05, 0.1), c(0.1, 0.2)) -->
<!-- medium_range <- cbind(c(0.075, 0), c(0, 0.2)) -->
<!-- long_range <- cbind(c(0.15, 0), c(0, 0.4)) -->
<!-- ``` -->

<!-- As in the previous sections, we sample from this distribution and compute the corresponding Papangelou conditional intensity for an individual of the first species. -->

<!-- ```{r} -->
<!-- plot_papangelou(model = model, -->
<!--                 medium_range_model = medium_range_model, -->
<!--                 alpha = alpha, -->
<!--                 beta0 = beta0, -->
<!--                 gamma = gamma, -->
<!--                 short_range = short_range, -->
<!--                 medium_range = medium_range, -->
<!--                 long_range = long_range) -->
<!-- ``` -->

<!-- It can be seen that the new individual exhibits repulsion from individuals of its own species (the circles) and is instead attracted to individuals of the other species (the triangles). -->
